---
title: "Earthquake Data Analysis"
subtitle: "Lake Tahoe Region - December 1948"
author: "Your Name"
date: today
format:
  revealjs:
    slide-number: true
    transition: slide
    background-transition: fade
    code-fold: true
    code-summary: "Show Code"
    theme: moon
    footer: "Earthquake Analysis Project"
---

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(knitr)

earthquake <- readRDS("data/processed/earthquake_clean.rds")
```

---

## Project Overview

- **Dataset:** Earthquake recordings from December 1948
- **Region:** Lake Tahoe area (California-Nevada border)
- **Observations:** `r nrow(earthquake)` earthquake events
- **Recording Stations:** `r n_distinct(earthquake$Location)` locations across `r n_distinct(earthquake$State)` states
- **Distance Range:** `r round(min(earthquake$Distance_km, na.rm=TRUE))` to `r round(max(earthquake$Distance_km, na.rm=TRUE))` km

---

## Research Questions

::: {.incremental}
- How are earthquake magnitudes distributed?
- What is the geographic pattern of recordings?
- How does earthquake intensity change with distance from epicenter?
- What are the correlations between magnitude, distance, and intensity?
:::

---

## Magnitude Distribution

```{r}
#| fig-width: 10
#| fig-height: 5

ggplot(earthquake, aes(x = Magnitude)) +
  geom_histogram(binwidth = 0.5, fill = "steelblue", color = "black", alpha = 0.7) +
  geom_vline(aes(xintercept = mean(Magnitude, na.rm = TRUE)), 
             color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Earthquake Magnitude Distribution",
    x = "Magnitude (Richter Scale)",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

**Mean Magnitude:** `r round(mean(earthquake$Magnitude, na.rm=TRUE), 2)` | **Range:** `r min(earthquake$Magnitude, na.rm=TRUE)` to `r max(earthquake$Magnitude, na.rm=TRUE)`

---

## Magnitude Summary Statistics

```{r}
mag_summary <- tibble(
  Metric = c("Mean", "Median", "Std Dev", "Min", "Max"),
  Value = c(
    round(mean(earthquake$Magnitude, na.rm = TRUE), 2),
    round(median(earthquake$Magnitude, na.rm = TRUE), 2),
    round(sd(earthquake$Magnitude, na.rm = TRUE), 2),
    min(earthquake$Magnitude, na.rm = TRUE),
    max(earthquake$Magnitude, na.rm = TRUE)
  )
)

kable(mag_summary)
```

---

## Recording Locations (Top 10)

```{r}
#| fig-width: 10
#| fig-height: 5

earthquake %>%
  count(Location, sort = TRUE) %>%
  head(10) %>%
  ggplot(aes(x = reorder(Location, n), y = n)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = n), hjust = -0.3, size = 3.5) +
  labs(
    title = "Top 10 Recording Locations",
    x = "Location",
    y = "Count"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14)) +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

---

## Geographic Coverage (Top States)

```{r}
state_summary <- earthquake %>%
  count(State, sort = TRUE)

# Show only top 10 states
state_summary %>% 
  head(10) %>%
  kable(col.names = c("State", "Observations"))
```

---

## Geographic Coverage (Additional States)

```{r}
# Show states 11-20
state_summary %>% 
  slice(11:20) %>%
  kable(col.names = c("State", "Observations"))
```

Total: `r n_distinct(earthquake$Location)` unique stations across `r n_distinct(earthquake$State)` states

---

## Distance vs Intensity Pattern

```{r}
#| fig-width: 10
#| fig-height: 5

intensity_summary <- earthquake %>%
  filter(!is.na(Intensity)) %>%
  group_by(Distance_Category) %>%
  summarise(
    Mean_Intensity = mean(Intensity, na.rm = TRUE), 
    Count = n(),
    .groups = "drop"
  ) %>%
  mutate(Distance_Category = factor(Distance_Category, 
                                     levels = c("Near", "Medium", "Far", "Very Far")))

ggplot(intensity_summary, aes(x = Distance_Category, y = Mean_Intensity)) +
  geom_col(fill = "steelblue", alpha = 0.7) +
  geom_text(aes(label = round(Mean_Intensity, 2)), vjust = -0.5, size = 4) +
  labs(
    title = "Average Intensity by Distance",
    x = "Distance from Epicenter",
    y = "Mean Intensity"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)))
```

---

## Magnitude by Distance

```{r}
#| fig-width: 10
#| fig-height: 5

earthquake %>%
  mutate(Distance_Category = factor(Distance_Category, 
                                     levels = c("Near", "Medium", "Far", "Very Far"))) %>%
  ggplot(aes(x = Distance_Category, y = Magnitude, fill = Distance_Category)) +
  geom_boxplot(alpha = 0.7) +
  labs(
    title = "Magnitude Range by Distance",
    x = "Distance Category",
    y = "Magnitude (Richter Scale)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14),
        legend.position = "none")
```

**Key Finding:** Near distances record smaller events; far distances captured the main shock (M8+)

---

## Intensity by Magnitude and Distance (Near & Medium)

```{r}
magnitude_intensity <- earthquake %>%
  filter(!is.na(Intensity)) %>%
  mutate(Magnitude_Group = cut(Magnitude, 
                               breaks = c(0, 7, 8, 9),
                               labels = c("M < 7", "M 7-8", "M > 8"))) %>%
  group_by(Distance_Category, Magnitude_Group) %>%
  summarise(
    Mean_Intensity = mean(Intensity, na.rm = TRUE),
    Count = n(),
    .groups = "drop"
  ) %>%
  mutate(Distance_Category = factor(Distance_Category, 
                                     levels = c("Near", "Medium", "Far", "Very Far")))

# Show only Near and Medium distances
magnitude_intensity %>% 
  filter(Distance_Category %in% c("Near", "Medium")) %>%
  kable(digits = 2)
```

---

## Intensity by Magnitude and Distance (Far & Very Far)

```{r}
# Show only Far and Very Far distances
magnitude_intensity %>% 
  filter(Distance_Category %in% c("Far", "Very Far")) %>%
  kable(digits = 2)
```

---

## Correlation Analysis - Magnitude vs Distance

```{r}
#| fig-width: 10
#| fig-height: 5

earthquake %>%
  ggplot(aes(x = Distance_km, y = Magnitude)) +
  geom_point(alpha = 0.5, color = "darkblue", size = 2) +
  geom_smooth(method = "lm", color = "red", linewidth = 1, se = TRUE) +
  labs(
    title = "Magnitude vs Distance Relationship",
    x = "Distance (km)",
    y = "Magnitude (Richter Scale)"
  ) +
  theme_minimal()
```

---

## Correlation Analysis - Magnitude vs Intensity

```{r}
#| fig-width: 10
#| fig-height: 5

earthquake %>%
  filter(!is.na(Intensity)) %>%
  ggplot(aes(x = Magnitude, y = Intensity)) +
  geom_point(alpha = 0.5, color = "darkgreen", size = 2) +
  geom_smooth(method = "lm", color = "red", linewidth = 1, se = TRUE) +
  labs(
    title = "Magnitude vs Intensity Relationship",
    x = "Magnitude",
    y = "Intensity (MMI)"
  ) +
  theme_minimal()
```

---

## Correlation Analysis - Distance vs Intensity

```{r}
#| fig-width: 10
#| fig-height: 5

earthquake %>%
  filter(!is.na(Intensity)) %>%
  ggplot(aes(x = Distance_km, y = Intensity, color = Magnitude)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", color = "black", linewidth = 1, se = TRUE) +
  scale_color_gradient(low = "blue", high = "red") +
  labs(
    title = "Distance vs Intensity (Seismic Attenuation)",
    x = "Distance (km)",
    y = "Intensity (MMI)",
    color = "Magnitude"
  ) +
  theme_minimal()
```

---

## Seismic Wave Attenuation

```{r}
distance_composition <- earthquake %>%
  mutate(Distance_Category = factor(Distance_Category, 
                                     levels = c("Near", "Medium", "Far", "Very Far"))) %>%
  group_by(Distance_Category) %>%
  summarise(
    Count = n(),
    Mean_Magnitude = round(mean(Magnitude, na.rm = TRUE), 2),
    Mean_Intensity = round(mean(Intensity, na.rm = TRUE), 2),
    .groups = "drop"
  )

kable(distance_composition)
```

---

## Summary Statistics

```{r}
summary_table <- tibble(
  Variable = c("Magnitude", "Distance (km)", "Intensity"),
  Mean = c(
    round(mean(earthquake$Magnitude, na.rm = TRUE), 2),
    round(mean(earthquake$Distance_km, na.rm = TRUE), 1),
    round(mean(earthquake$Intensity, na.rm = TRUE), 2)
  ),
  Min = c(
    min(earthquake$Magnitude, na.rm = TRUE),
    round(min(earthquake$Distance_km, na.rm = TRUE), 1),
    min(earthquake$Intensity, na.rm = TRUE)
  ),
  Max = c(
    max(earthquake$Magnitude, na.rm = TRUE),
    round(max(earthquake$Distance_km, na.rm = TRUE), 1),
    max(earthquake$Intensity, na.rm = TRUE)
  )
)

kable(summary_table)
```

---

## Key Insights from Analysis

:::{.incremental}
- **Magnitude Distribution:** Main shock at M8 with foreshocks/aftershocks
- **Geographic Pattern:** Dense coverage across CA, NV, and neighboring states
- **Physical Attenuation:** Intensity decreases with distance from epicenter
- **Data Quality:** Multiple magnitude earthquakes at different distances
- **Seismic Physics:** Correlations confirm known seismic wave propagation patterns
:::

---

## Conclusions

- **Main Event:** Magnitude 8 earthquake on December 16, 1948 in Lake Tahoe region
- **Extensive Network:** 200+ observations recorded across multiple states
- **Physics Confirmed:** Seismic waves lose intensity with distance
- **Data Quality:** High-quality recording of a major earthquake event
- **Research Value:** Valuable dataset for studying seismic wave propagation

---

## Thank You

Questions?
